\texttt{thornado}-transport: IMEX schemes for two-moment neutrino transport respecting Fermi-Dirac statistics

### 2019-03-19 Documents/M1/Astronum_2018/sections/Numerical.tex

In this section we consider a more ``realistic'' test: two-dimensional multigroup neutrino transport with emission, absorption, and isoenergetic scattering through a stationary background.  
This test is designed to test the realizability-preserving properties of the PD-ARS schemes.  
Figure~\ref{fig:NeutrinoStationaryTestEOS} plots the thermal state and the opacities used (adopted from~\cite{Bruenn_1985}).  
This test is computed on a two-dimensional domain $D=\{\vect{x}\in\bbR^{2}:x^{1}\in[0,200], x^{2}\in[0,200]\}$ with a grid of 128 elements in each direction, a reflecting inner boundary and a homogeneous out-flowing outer boundary.
We use 10 energy groups to cover the range from 0~MeV to 300~MeV and initialize the neutrino number density to $\cJ = 10^{-99}$, flux density to $\bcH=0$, and evolve until an approximate steady state is reached ($t=5$~ms).  
For this test, we employed both CB and Minerbo closures.  
We attempted to run this test with our PD-ARS schemes, SSP2332 from~\cite{pareschiRusso_2005}, IMEXRKCB2 proposed by Cavaglieri \& Bewley~\cite{cavaglieriBewley2015}, and the IMEX PC2 scheme proposed by McClarren et al.~\cite{mcclarren_etal_2008}.
Only the PD-ARS schemes produce realizable moments and are able to evolve to a steady state with either CB or Minerbo closure.
SSP2332 and IMEX PC2 failed after a few time steps with either CB or Minerbo closure because of the development of unrealizable moments.
Even though IMEXRKCB2 with Minerbo closure can run and reach a steady state in this test, its results are different from that of PD-ARS schemes with CB closure, and there is no guarantee of stability.
%It's difficult to determine whether the results are in fact compromised by a particular correction step and, if so, in what way and to what degree.

>>>>

In this section we consider a more ``realistic'' test: two-dimensional multigroup neutrino transport with emission, absorption, and isoenergetic scattering through a stationary background.  
This test is designed to test the realizability-preserving properties of the PD-ARS schemes.  
Figure~\ref{fig:NeutrinoStationaryTestEOS} left plots the initial thermal state: 
\begin{subequations}
\begin{align}
\text{Density: }~\rho & = 4 \times 10^{14} \times \dfrac{7.5}{( 7.5 + ( r / 5 )^4 )} ~\text{g $\cdot$ cm$^{-3}$}\\
\text{Temperature: }~T & = 1.5 \times 10^{11} \times \dfrac{1}{1 + ( r / 50 )^2 } ~\text{K}\\
\text{Electron Fraction: }~Y_e & = 0.25 \times \left(  1 + \dfrac{1}{1+ ( r / 50 )^{-12}}  \right) 
\end{align}
\label{eq:NuStatinaryStateEOS}
\end{subequations}
with radius $r$ in kilometer unit. 
Eq.~\eqref{eq:NuStatinaryStateEOS} is designed to simulate a CCSN progenitor.
Figure~\ref{fig:NeutrinoStationaryTestEOS} right plots the used opacities which are provided by interpolations of a pre-calculated opacity table based on~\cite{Bruenn_1985}.
This test is computed on Cartesian coordinate with a two-dimensional domain $D=\{\vect{x}\in\bbR^{2}:x^{1}\in[0,200], x^{2}\in[0,200]\}$ in kilometer unit, a grid of 128 elements in each direction, 10 energy groups covering $(0,300)$~MeV, a reflecting inner boundary and a homogeneous out-flowing outer boundary.
Because of the Cartesian coordinate, we ignored some geometric terms, which leads the initial condition to be non-axisymmetric.
We initialize the neutrino number density to $\cJ = 10^{-99}$, flux density to $\bcH=0$, and evolve until an approximate steady state is reached ($t=5$~ms).
No hydrodynamics is used for the stationary background.
For this test, we employed both CB and Minerbo closures.  
We attempted to run this test with our PD-ARS schemes, SSP2332 from~\cite{pareschiRusso_2005}, IMEXRKCB2 proposed by Cavaglieri \& Bewley~\cite{cavaglieriBewley2015}, and the IMEX PC2 scheme proposed by McClarren et al.~\cite{mcclarren_etal_2008}.
Only the PD-ARS schemes produce realizable moments and are able to evolve to a steady state with either CB or Minerbo closure.
SSP2332 and IMEX PC2 failed after a few time steps with either CB or Minerbo closure because of the development of unrealizable moments.
Even though IMEXRKCB2 with Minerbo closure can run and reach a steady state in this test, its results are different from that of PD-ARS schemes with CB closure, and there is no guarantee of stability.
%It's difficult to determine whether the results are in fact compromised by a particular correction step and, if so, in what way and to what degree.

### 2019-03-19 Documents/M1/Astronum_2018/sections/TimeIntegration.tex

Treating the transport operator explicitly and the collision operator implicitly, a standard $s$-stage IMEX scheme takes the following form~\cite{pareschiRusso_2005}: 
\begin{align}
  \vect{u}^{(i)}
  &=\vect{u}^{n}
  +\dt\sum_{j=1}^{i-1}\tilde{a}_{ij}\,\vect{\cT}(\vect{u}^{(j)})
  +\dt\sum_{j=1}^{i}a_{ij}\,\vect{\cQ}(\vect{u}^{(j)}),
  \quad i=1,\ldots,s, \label{imexStages} \\
  \vect{u}^{n+1}
  &=\vect{u}^{n}
  +\dt\sum_{i=1}^{s}\tilde{w}_{i}\,\vect{\cT}(\vect{u}^{(i)})
  +\dt\sum_{i=1}^{s}w_{i}\,\vect{\cQ}(\vect{u}^{(i)}), \label{imexIntermediate} 
\end{align}
where $(\tilde{a}_{ij})$ and $(a_{ij})$, coefficients of the $i$-th stage, are elements of matrices $\tilde{A}$ and $A$, respectively.
The matrices $\tilde{A}$ and $A$ are lower triangular ($\tilde{A}$ is strictly lower triangular so that the transport part is explicit).  
The vectors $\tilde{\vect{w}}=(\tilde{w}_{1},\ldots,\tilde{w}_{s})^{T}$ and $\vect{w}=(w_{1},\ldots,w_{s})^{T}$ are the weights in the assembly step in Eq.~\eqref{imexIntermediate}.
These coefficients and weights must satisfy certain order conditions for consistency, accuracy, and other properties.  
For second-order temporal accuracy, the following conditions are required~\cite{hairer_1981}:
\begin{equation}
  \sum_{i=1}^{s}\tilde{w}_{i}=\sum_{i=1}^{s}w_{i}=1,
  \label{orderConditions1}
\end{equation}
and
\begin{equation}
  \sum_{i=1}^{s}\tilde{w}_{i}\,\tilde{c}_{i}
  =\sum_{i=1}^{s}\tilde{w}_{i}\,c_{i}
  =\sum_{i=1}^{s}w_{i}\,\tilde{c}_{i}
  =\sum_{i=1}^{s}w_{i}\,c_{i}=\f{1}{2}, 
  \label{orderConditions2}
\end{equation}
where $\tilde{c}_{i} = \sum_{j=1}^{s}\tilde{a}_{ij}$ and $c_{i}=\sum_{j=1}^{s}a_{ij}$.

>>>>

Treating the transport operator explicitly and the collision operator implicitly, a standard $s$-stage IMEX scheme takes the following form~\cite{pareschiRusso_2005}: 
\begin{align}
  \vect{u}^{(i)}
  &=\vect{u}^{n}
  +\dt\sum_{j=1}^{i-1}\tilde{a}_{ij}\,\vect{\cT}(\vect{u}^{(j)})
  +\dt\sum_{j=1}^{i}a_{ij}\,\vect{\cQ}(\vect{u}^{(j)}),
  \quad i=1,\ldots,s, \label{imexStages} \\
  \vect{u}^{n+1}
  &=\vect{u}^{n}
  +\dt\sum_{i=1}^{s}\tilde{w}_{i}\,\vect{\cT}(\vect{u}^{(i)})
  +\dt\sum_{i=1}^{s}w_{i}\,\vect{\cQ}(\vect{u}^{(i)}), \label{imexIntermediate} 
\end{align}
where $(\tilde{a}_{ij})$ and $(a_{ij})$, coefficients of the $i$-th stage, are elements of matrices $\tilde{A}$ and $A$, respectively.
The matrices $\tilde{A}$ and $A$ are lower triangular.
($\tilde{A}$ is strictly lower triangular so that the transport part is explicit.)  
The vectors $\tilde{\vect{w}}=(\tilde{w}_{1},\ldots,\tilde{w}_{s})^{T}$ and $\vect{w}=(w_{1},\ldots,w_{s})^{T}$ are the weights in the assembly step in Eq.~\eqref{imexIntermediate}.
These coefficients and weights must satisfy certain order conditions for consistency, accuracy, and other properties.  
For second-order temporal accuracy, the following conditions are required~\cite{hairer_1981}:
\begin{equation}
  \sum_{i=1}^{s}\tilde{w}_{i}=\sum_{i=1}^{s}w_{i}=1,
  \label{orderConditions1}
\end{equation}
and
\begin{equation}
  \sum_{i=1}^{s}\tilde{w}_{i}\,\tilde{c}_{i}
  =\sum_{i=1}^{s}\tilde{w}_{i}\,c_{i}
  =\sum_{i=1}^{s}w_{i}\,\tilde{c}_{i}
  =\sum_{i=1}^{s}w_{i}\,c_{i}=\f{1}{2}, 
  \label{orderConditions2}
\end{equation}
where $\tilde{c}_{i} = \sum_{j=1}^{s}\tilde{a}_{ij}$ and $c_{i}=\sum_{j=1}^{s}a_{ij}$.

### 2019-03-19 Documents/M1/Astronum_2018/sections/TwoMoments.tex

\subsection{Transport Equations}
We consider neutrino transport through a static background, and include neutrino-matter interactions due to emission, absorption, and isoenergetic scattering.
After scaling to dimensionless units, the Boltzmann equation can be written as
\begin{equation}
  \pd{f}{t}+\vect{\ell}\cdot\nabla f
  =\f{1}{\tau}\,\cC(f),
  \label{eq:boltzmann}
\end{equation}
where the distribution function $f = f(\omega,\varepsilon,\vect{x},t)$ gives the number of neutrinos propagating in the direction $\omega\in\bbS^{2}$, with energy $\varepsilon\in\bbR^{+}$, at position $\vect{x}\in\bbR^{3}$ and time $t\in\bbR^{+}$.  
$\vect{\ell}(\omega)\in\bbR^{3}$ is the unit vector parallel to the neutrino three-momentum: $\vect{p}=\varepsilon\,\vect{\ell}$.
On the right-hand side, $\tau$ is a collision time scale.  
In opaque regions, where neutrinos have frequent interactions with the background, $\tau\ll1$.  
In transparent regions, where neutrinos rarely interact and stream freely, $\tau\gg1$.
The collision term, $\cC(f)$, which models emission, absorption, and isoenergetic scattering is given by
\begin{equation}
  \cC(f)=\xi\,\big(\,f_{0}-f\,\big)
  +(1-\xi)\,\big(\,\f{1}{4\pi}\int_{\bbS^{2}}f\,d\omega-f\,\big),
  \label{eq:collisionTerm}
\end{equation}
where $\xi = \sigma_{\Ab} / (\sigma_{\Ab}  + \sigma_{\Scatt} )$ is the ratio of the absorption opacity $\sigma_{\Ab}$ to the total opacity.  
The scattering opacity is $\sigma_{\Scatt}$.  
The limit $\xi = 1$, when $\sigma_{\Scatt} = 0$, corresponds to pure absorption, while $\xi = 0$, when $\sigma_{\Ab} = 0$, corresponds to pure scattering.  
The equilibrium distribution function for neutrinos is given by the Fermi-Dirac distribution:
\begin{equation}
  f_{0}(\vect{z})=\f{1}{e^{(\varepsilon-\mu(\vect{x}))/T(\vect{x})}+1}, 
  \label{eq:fermiDirac}
\end{equation}
where $\vect{z}:=\{\varepsilon,\vect{x}\}$, $T$ is the material temperature in energy units and $\mu$ is the neutrino chemical potential.
Both $T$ and $\mu$ depend on the spatial position $\vect{x}$.

@@

Neutrinos are fermions and obey the Pauli exclusion principle.  
Because of this, the neutrino distribution function is bounded; i.e.~$f \in [0,1]$.
As a result, the angular moments $\cJ$ and $\vect{\cH}$ and the Eddington factor $\chi$ satisfy the following bounds~\cite{levermore_1984,lareckiBanach_2011,kershaw_1976,shohatTamarkin_1943}: 
\begin{align}
\cJ \in[0,1], \quad &(1-\cJ)\cJ-|\vect{\cH}| \geq 0, \label{eq:MomentsBounds} \\
  \chi_{\mbox{\tiny min}}
  =\max\big(1-\f{2}{3\cJ},h^{2}\big)
  \leq & \chi \leq \min\big(1,\f{1}{3\cJ}-\f{\cJ}{1-\cJ}h^{2}\big)=\chi_{\mbox{\tiny max}},
  \label{eq:eddingtonFactorBounds}
\end{align}
where $h = |\bcH|/\cJ$ is the flux factor.

>>>>

\subsection{Transport Equations}
We consider neutrino transport through a static background under the Newtonian framework.
We include neutrino-matter interactions due to emission, absorption, and isoenergetic scattering, which dominate the opacity under the physical conditions that we're interested in.
It's possible to include other interactions, such as neutrino-electron scattering and neutrino pair production, in the two-moment model.
But involving those interactions complicates the collision term and requires more efforts in proving the realizability of a scheme.
We leave it for future work.
After scaling to dimensionless units, the Boltzmann equation can be written as
\begin{equation}
  \pd{f}{t}+\vect{\ell}\cdot\nabla f
  =\f{1}{\tau}\,\cC(f),
  \label{eq:boltzmann}
\end{equation}
where the distribution function $f = f(\omega,\varepsilon,\vect{x},t)$ gives the number of neutrinos propagating in the direction $\omega\in\bbS^{2}$, with energy $\varepsilon\in\bbR^{+}$, at position $\vect{x}\in\bbR^{3}$ and time $t\in\bbR^{+}$.  
$\vect{\ell}(\omega)\in\bbR^{3}$ is the unit vector parallel to the neutrino three-momentum: $\vect{p}=\varepsilon\,\vect{\ell}$.
On the right-hand side, $\tau$ is a collision time scale.  
In opaque regions, where neutrinos have frequent interactions with the background, $\tau\ll1$.  
In transparent regions, where neutrinos rarely interact and stream freely, $\tau\gg1$.
The collision term, $\cC(f)$, which models emission, absorption, and isoenergetic scattering is given by
\begin{equation}
  \cC(f)=\xi\,\big(\,f_{0}-f\,\big)
  +(1-\xi)\,\big(\,\f{1}{4\pi}\int_{\bbS^{2}}f\,d\omega-f\,\big),
  \label{eq:collisionTerm}
\end{equation}
where $\xi = \sigma_{\Ab} / (\sigma_{\Ab}  + \sigma_{\Scatt} )$ is the ratio of the absorption opacity $\sigma_{\Ab}$ to the total opacity.  
The scattering opacity is $\sigma_{\Scatt}$.  
The limit $\xi = 1$, when $\sigma_{\Scatt} = 0$, corresponds to pure absorption, while $\xi = 0$, when $\sigma_{\Ab} = 0$, corresponds to pure scattering.  
The equilibrium distribution function for neutrinos is given by the Fermi-Dirac distribution:
\begin{equation}
  f_{0}(\vect{z})=\f{1}{e^{(\varepsilon-\mu(\vect{x}))/T(\vect{x})}+1}, 
  \label{eq:fermiDirac}
\end{equation}
where $\vect{z}:=\{\varepsilon,\vect{x}\}$, $T$ is the material temperature in energy units and $\mu$ is the neutrino chemical potential.
Both $T$ and $\mu$ depend on the spatial position $\vect{x}$.

@@

Neutrinos are fermions and obey the Pauli exclusion principle.  
Because of this, the neutrino distribution function is bounded; i.e.~$f \in [0,1]$.
As a result, the angular moments $\cJ$ and $\vect{\cH}$ and the Eddington factor $\chi$ satisfy the following bounds~\cite{levermore_1984,lareckiBanach_2011,kershaw_1976,shohatTamarkin_1943}: 
\begin{align}
\cJ \in[0,1], \quad &(1-\cJ)\cJ-|\vect{\cH}| \geq 0, \label{eq:MomentsBounds} \\
  \chi_{\mbox{\tiny min}}
  =\max\big(1-\f{2}{3\cJ},h^{2}\big)
  \leq & \chi \leq \min\big(1,\f{1}{3\cJ}-\f{\cJ}{1-\cJ}h^{2}\big)=\chi_{\mbox{\tiny max}},
  \label{eq:eddingtonFactorBounds}
\end{align}
where $h = |\bcH|/\cJ$ is the flux factor.
These relationships hold for moment in the comoving frame.
Similar relationships can be derived in the laboratory frame, which we leave for future work.

### 2019-03-19 Documents/M1/Astronum_2018/sections/intro.tex

Applying the two-moment method does simplify the problem, but doesn't guarantee an affordable solution.
To be precise, how to discretize the continuous system of equation given by the two-moment method and solve the discretized system efficiently remains a question.
In fact, the time scales of neutrino interactions with the background (they can be $\sim\mathcal{O}(10^{-13})$~second) is short compared to the duration of the CCSN explosion ($\sim\mathcal{O}(1)$~second).  
This means that $\sim\mathcal{O}(10^{13})$ time steps could be needed for solving the system fully explicitly. 
On the other hand, solving the moment equations fully implicitly requires inverting global band-structured matrices whose sizes depend on the phase-space discretization.
Such a global inversion is both expensive and unfriendly to parallelization.
To circumvent these challenges, implicit-explicit (IMEX) methods are taken into consideration.
By treating the transport terms in the two-moment equations explicitly and the collision terms implicitly, IMEX methods are subject only to a time step governed by the explicit transport terms, and the matrices to be inverted are block diagonal.
Therefore, IMEX methods require far fewer time steps compared with a fully explicit method, and the computation for each step is easily parallelizable.
For the relativistic setting that we have, where the fluid and the neutrinos have comparable propagation speeds, IMEX methods can be efficient.

>>>>

Applying the two-moment method does simplify the problem, but doesn't guarantee an affordable solution.
To be precise, how to discretize the continuous system of equatio
ns given by the two-moment method and solve the discretized system efficiently remains a question.
In fact, the time scales of neutrino interactions with the background (they can be $\sim\mathcal{O}(10^{-13})$~second) is short compared to the duration of the CCSN explosion ($\sim\mathcal{O}(1)$~second).  
This means that $\sim\mathcal{O}(10^{13})$ time steps could be needed for solving the system fully explicitly. 
On the other hand, solving the moment equations fully implicitly requires inverting global band-structured matrices whose sizes depend on the phase-space discretization.
Such a global inversion is both expensive and unfriendly to parallelization.
To circumvent these challenges, implicit-explicit (IMEX) methods are taken into consideration.
By treating the transport terms in the two-moment equations explicitly and the collision terms implicitly, IMEX methods are subject only to a time step governed by the explicit transport terms, and the matrices to be inverted are block diagonal.
Therefore, IMEX methods require far fewer time steps compared with a fully explicit method, and the computation for each step is easily parallelizable.
For the relativistic setting that we have, where the fluid and the neutrinos have comparable propagation speeds, IMEX methods can be efficient.
